Longitudinal Manifold Estimation

### 2023-04-25 paper/main.tex
initial algorithm updates

At each time point, the estimated principal manifold has the functional form of a smoothing spline or thin plate spline, depending on the intrinsic dimension $d$. As detailed by \cite{greenSilverman1994}, smoothing splines, and thin plate splines as their higher dimensional analogues, may be considered to have knots at the points being used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fit must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\mathbf{r}_i\right\}_{i=1}^{N}$ ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions for each time point $f_t$. This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing
\[%
  \mathcal{L}(f_t) = (\mathbf{X_t} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t)^{T}(\mathbf{X} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t) + \gamma\mathbf{s}_t^{T}E\mathbf{s}_t
,\]%
where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $R$ has $R_{ij} = \phi_i(\mathbf{r}_j)$, where $\phi_i(\cdot)$ is a linear basis function. This minimization is conducted by solving
\[%
  \left[
    \begin{array}{cc}
      E + \lambda_t^* I & R^{T} \\
      R^{T} & 0
    \end{array}
  \right]\left(
    \begin{array}{c}
      \mathbf{s}_t \\
      \mathbf{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \mathbf{X}_t \\
      0
    \end{array}
  \right)
,\]%
yielding function
\[%
  f_t^*(r) = \sum_{j=1}^{N}s_{jt}^* \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{kt}^*\phi(r)_k
.\]%

@@

Once longitudinal estimates $f_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross-validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows:
\[%
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|Y_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(Y_{i, t}))\|^2
.\]%

The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.

@@

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
\KwData{Data points $\left\{y_{i, t}\right\}_{i, t}^{I_t, T}$, positive integer $d$, positive integer $N_0 < I - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $\hat{g}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, \dots, T$} {
  Apply HDMDE algorithm with input $\left(\left\{y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, N_t, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, N_t, t}\right\}_{j = 1}^{N_t}$\;
}

Parameterization: Apply ISOMAP to parameterize $\left\{\mu_{j, N_t, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$. Formally set $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets r_j$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;

\For{t = 1, 2, \dots, T} {
  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t^*$ and $\lambda_t^*$\;
  $\left\{r_{j, t}\right\} \gets \pi_{f_t^*}(\mu_{j, N_t, t})$ for $j = 1, \dots, N_t$ and $x_{j, t} \gets f_t^*(r_{j, t})$ for $j = 1, \dots, N_t$\;
}

Let $N = \sum_{t=1}^{T}N_t$, $\mathbf{x}$ be a $N \times D$ matrix of stacked values of $\left\{x_{j, t}\right\}_{t = 1}^T$, and $\mathbf{r}$ be a $N \times d + 1$ matrix of stacked values of $\left\{r_{j, t}\right\}_{t = 1}^T$ \;

@@

Questions:
\begin{enumerate}
  \item What is the best way to evaluate the performance of the estimates? We can use SSD for the embedded values, but how to assess the parameterized values?
  \item Should there be updates to what is defined as the "optimal" model in the PME algorithm? Currently it is possible for a model with non-minimal SSD to be chosen within the fitting of a single tuning value.
  \item Why would we see the SSD increase substantially in late optimization steps?
  \item Generally, what are the limitations on the estimated manifolds intersecting? How close can it come to intersecting without showing strange behavior?
  \item Is thin-plate spline computational strategy from Wahba the best approach, or could updated computational approach help (Wood, 2003)?
  \item How much error does the registration process introduce? Can the LPME algorithm successfully handle a certain number of bad registrations or extractions?
\end{enumerate}

>>>>

At each time point, the estimated principal manifold has the functional form of a smoothing spline or thin plate spline, depending on the intrinsic dimension $d$. As detailed by \cite{greenSilverman1994}, smoothing splines, and thin plate splines as their higher dimensional analogues, may be considered to have knots at the points being used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fit must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\mathbf{r}_i\right\}_{i=1}^{N}$ ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions for each time point $f_t$. This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing
\[%
  \mathcal{L}(f_t) = (\mathbf{X_t} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t)^{T}(\mathbf{X} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t) + \gamma\mathbf{s}_t^{T}E\mathbf{s}_t
,\]%
where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $R$ has $R_{ij} = \phi_i(\mathbf{r}_j)$, where $\phi_i(\cdot)$ is a linear basis function. This minimization is conducted by solving
\[%
  \left(
    \begin{array}{cc}
      E + \lambda_t^* I & R^{T} \\
      R^{T} & 0
    \end{array}
  \right)\left(
    \begin{array}{c}
      \mathbf{s}_t \\
      \mathbf{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \mathbf{X}_t \\
      0
    \end{array}
  \right)
,\]%
yielding function
\[%
  f_t^*(r) = \sum_{j=1}^{N}s_{jt}^* \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{kt}^*\phi(r)_k
.\]%

@@

Once longitudinal estimates $f_{\gamma}(t, \mathbf{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows:
\[%
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|Y_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(Y_{i, t}))\|^2
.\]%
The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$.

Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.

@@

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
  \KwData{Data points $\left\{Y_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, \dots, T$} {
  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
}

  Apply ISOMAP to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$. Formally set $\pi_{f_t(0)}(\mu_{j, t}) \gets r_{j, t}$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;

\For{t = 1, 2, \dots, T} {
  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$ and $\lambda_t^*$\;
  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
}

  Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$, and $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$ and $t = 1, \dots, T$\;
Let $N = \sum_{t=1}^{T}N_t$, $\mathbf{x}$ be a $N \times D$ matrix of stacked values of $\left\{x_{j, t}\right\}_{t = 1}^T$, and $\mathbf{r}$ be a $N \times d + 1$ matrix of stacked values of $\left\{r_{j, t}\right\}_{t = 1}^T$ \;

@@
